---
title: "Part2: Toxicity Analysis"
runtime: shiny
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(warn=-1)
library(shiny)
library(dplyr)
require(gridExtra)
library(ggplot2)
library(reshape2)
library(tidyverse)
library(shinydashboard)
library(DT)
    
```
## Overall Analysis
- Lethal dose (LD50) is the amount of an ingested substance that kills 50 percent of a test sample
- The LD50 values in the following slides are the oral toxicity levels for a rat. 
- The Veg1 dataset shows that restricted use chemicals are often applied to household vegetables such as 
  cauliflower and broccoli above the standard LD50 rate
- As portrayed in the following slides, there is a upward trend of larger doses of pesticides being and greater 
variety of chemicals used in produce as the years go on as seen in the deviating t statistics

## Toxic Chemicals Found

The following is a comprehensive list of toxic (festricted use chemicals) found within broccoli 
and cauliflower across the US. 

```{r}
chemicals <- c('BETA-CYFLUTHRIN','BIFENTHRIN','CHLORANTRANILIPROLE','CHLORPYRIFOS','CYFLUTHRIN',
               'ESFENVALERATE','IMIDACLOPRID','LAMBDA-CYHALOTHRIN','METHOMYL','NALED',
               'PERMETHRIN','THIAMETHOXAM','ZETA-CYPERMETHRIN',
               'DIAZINON','DISULFOTON','CYPERMETHRIN','PRONAMIDE','GAMMA-CYHALOTHRIN')
               
toxicityLevels <- c(8350,70,5000,270,1271,458,450,79,48,281,4000,1563,
                    250,400,12.5,250,8350,79)
toxicity <- data.frame(chemicals,toxicityLevels)

renderDT({
    toxicity1 <- toxicity %>% slice(1:7) 
    toxicity1
  })
```

## Toxic Chemicals Found
```{r}
renderDT({
    toxicity2 <- toxicity %>% slice(7:14) 
    toxicity2
  })
```

## Toxic Chemicals Found
```{r}
renderDT({
    toxicity3 <- toxicity %>% slice(14:21) 
    toxicity3
  })
```

## Pesticide Yearly Trend
- The time series in the following slide shows the trend of toxic pesticide use on broccoli and cauliflower
- In many cases the amount of pesicide being used was within range or greater than the LD50 value that killed 
50% of rats - therefore harmful to us as well. 
- Chemicals found that were above standardized LD50 values include: Bifenthrin, Esfenvalerate, Imidaclorpid, Lambda-Cyhalothrin, Methomyl, Naled, 
Zeta-Cypermethrin and Cypermethrin

## Pesticide Yearly Trend
```{r eruptions}
    #only load data once 
    #read in excel data 
    veg.1 <- readxl::read_excel("veg1.xlsx")
    cnames.1 <- colnames(veg.1)
    c <- apply(veg.1, 2, n_distinct)
    #c[c>1]
    
    d <- names(c[c==1])
    e <- names(c[c>1])
    
    veg.2 <- select(veg.1, e)
    cnames.2 <- colnames(veg.2)
    
    #Veg.2 is filtered table w/o NA values 
    #apply(veg.2, 2, n_distinct)
    
    veg.3 <- dplyr::rename(veg.2, 
                           Geo = 'Geo Level', 
                           State = 'State ANSI',
                           Data = 'Data Item',
                           Category = 'Domain Category')
    
    #Veg.3 is w/ shortend column names
    cnames.3 <- colnames(veg.3)
    yy <- separate(veg.3, Category, into = c("label", "quant"), sep=",")

    #ru is table w/ data for only restricted use chemicals 
    ru <- filter(yy, label=="RESTRICTED USE CHEMICAL")
    ru2 <- filter(yy, label=="RESTRICTED USE CHEMICAL") 

    ru2 <- ru2%>%mutate(chem = str_extract(quant, "(?<=\\().*?(?= =)"))
    ru3 <- ru2%>%mutate(chem = str_extract(quant, "(?<=\\().*?(?= =)"))
    ru4 <- ru2%>%mutate(chem = str_extract(quant, "(?<=\\().*?(?= =)"))
    #View(ru4)
    
    #ru1 is table w/ types of diff restricted chemicals
    ru1 <- ru %>% select(label, quant) %>% unique()
    toxicity <- data.frame(chemicals,toxicityLevels)
    
inputPanel(
        title = "Time Series Controls",solidHeader = TRUE,
        sliderInput("years", "Years:",
                    min = 2006, max = 2016, value = c(2006,2016),sep = ""),

        selectInput("chemical", "Chemical:",chemicals), 
        selectInput("veggies1", "Vegetables:",c('BROCCOLI','CAULIFLOWER'))
)

renderPlot({
    #filter ru table by year and chemical type 
    test <- ru%>%filter(Commodity == input$veggies1)
    newru <- ru%>%filter(between(Year, input$years[1],input$years[2]), str_detect(quant, input$chemical), Commodity == input$veggies1)
    
    if(newru$Value == '(D)'){
    newru$Value[newru$Value == '(D)'] <- 0
    
    maxAmount <- toxicity%>%filter(chemicals == input$chemical)
    Amount <- ts(newru$Value, start = c(input$years[1]), end = input$years[2])
    plot.ts(Amount)
    title(main= paste(input$chemical, " IN ", input$veggies1, " - DISSOLVED (D)"))
    mtext(paste0("Max LD50 value: ", maxAmount$toxicityLevels), side = 3)
    }
   
    else{
    #add ld50 line 
    maxAmount <- toxicity%>%filter(chemicals == input$chemical)
    Amount <- ts(newru$Value, start = c(input$years[1]), end = input$years[2])
    plot.ts(Amount)
    title(main=input$chemical)
    mtext(paste0("Max LD50 value: ", maxAmount$toxicityLevels), side = 3)
    abline(h=maxAmount$toxicityLevels, col="blue")
    legend("topright", "Max LD50 value", col=c("blue"), lty=1, cex=.65)
    }
    
  })
```

## Average Toxicity Levels By Year
```{r}
inputPanel(
        title = "Bar Plot Controls", status = "danger",
        selectInput("years2", "Year:", c(2006,2010,2014,2016)),
        selectInput("veggies2", "Vegetable:", c('BROCCOLI', 'CAULIFLOWER'))
)

renderPlot({
    newru2 <- ru2%>%filter(Year == input$years2, Commodity == input$veggies2)
    #replace (D) with 0
    newru2$Value[newru2$Value == '(D)'] <- 0
    newru2$Value[newru2$Value == '(Z)'] <- 0

    chems <- newru2$chem
    vals <- newru2$Value
    print(chems)
    df1 = data.frame(chems, vals)
    
    test3 <- newru2 %>% 
      group_by(chem) %>% 
      summarise(mean_val = mean(as.numeric(Value)))
    print(test3)
    par(mar=c(4,11,4,2))
    barplot(test3$mean_val, names.arg = test3$chem, horiz = T, las = 2, col = "blue", xlab = "Amount (mg/kg)")
    title(main=paste(input$veggies2, " PESTICIDES IN " , input$years2))
  })
```

## T-Test
- Null Hypothesis: There is an increasing of variety and amount of chemicals applied to produce as years progress
- Conclusion:There was a significant difference in the paired t statistic and mean when comparing earlier years with more 
recent years. The results suggest that the null hypothesis is true and there is a trend of greater variety of chemicals with 
greater amounts. 

## T-Test
```{r}
inputPanel(
        title = "T-Test Controls", status = "danger",
        selectInput("yeartt", "Year:", c(2006,2010,2014,2016),selected = 2010),
        selectInput("yeartt2", "Year:", c(2006,2010,2014,2016), selected=2014), 
        selectInput("chemicals3", "Chemicals:", chemicals)
)

renderUI({
    y1 <- ru3%>%filter(Year == input$yeartt, chem == input$chemicals3)
    y2 <- ru4%>%filter(Year == input$yeartt2, chem == input$chemicals3)
    
    #replace values with NA so they tables are same length
    y1$Value[y1$Value == '(D)'] <- 0
    y2$Value[y2$Value == '(D)'] <- 0

    #conduct paired t test 
    s1 <- paste0(t.test(as.numeric(y1$Value),as.numeric(y2$Value),paired=TRUE))
    print(s1)
    print(t.test(as.numeric(y1$Value),as.numeric(y2$Value),paired=TRUE))
    HTML(paste0(
                "<ul><li>T statistic: ",s1[1],"</li>",
                "<li>95% confidence interval: ",s1[4],"</li>
                <li>Degrees of freedom: ",s1[2],"</li>
                <li>P-value: ",s1[3],"</li>
                <li>Mean of differences: ",s1[5],"</li><ul>"
                ))
  })

```



